Normalization by Evaluation for Call-by-Push-Value and Polarized Lambda-Calculus

### 2019-02-11 fscd19/fscd19.tex
[ fscd19 ] reflection and reification for cbpv

Simple types shall be distinguished into positive types $P$ and
negative types $N$, depending on their root type former.  Function
($\To$) and product types ($\times$ and $\tyone$) are negative,
while base types ($\tybase$) and sum types ($+$ and $\tyzero$) are positive.
\[
\begin{array}{lrl@{\qquad}l}
  A,B,C & ::= & P \mid N & \mbox{simple types} \\
  P     & ::= & \tyzero \mid A + B \mid \tybase & \mbox{positive types} \\
  N     & ::= & \tyone \mid A \times B \mid A \To B
  % P     & ::= & a \mid A + B \mid \tyzero & \mbox{positive types} \\
  % N     & ::= & A \To B \mid A \times B \mid \tyone
    & \mbox{negative types} \\
\end{array}
\]
% Simple types shall be given by the grammar
% % \fbox{$S,T ::= P \mid S \To T \mid S \times T \mid \tyone \mid S + T
% %   \mid \tyzero$}.
% \fbox{$A,B,C ::= a \mid A \To B \mid A \times B \mid \tyone \mid A + B
%   \mid \tyzero$}.
% Herein, $a$ ranges over atomic types aka base types.
Intrinsically well-typed lambda-terms, in abstract syntax,
are just inhabitants $t$ of the indexed set \fbox{$A \from \Gamma$},
inductively defined by the following rules.
\begin{gather*}
  \nru{\tvar}{A \in \Gamma}{A \from \Gamma}
\qquad
  \nru{\tabs}{B \from \cext \Gamma A}{A \To B \from \Gamma}
\qquad
  \nru{\tapp}
    {A \To B \from \Gamma \qquad A \from \Gamma}
    {B \from \Gamma}
\\[2ex]
  \nru{\tunit}{}{\tyone \from \Gamma}
\qquad
  \nru{\tpair}
    {A_1 \from \Gamma \qquad A_2 \from \Gamma}
    {A_1 \times A_2 \from \Gamma}
\qquad
  \nru{\tprj_i}
    {A_1 \times A_2 \from \Gamma}
    {A_i \from \Gamma}
\\[2ex]
  \nru{\tinj_i}
    {A_i \from \Gamma}
    {A_1 + A_2 \from \Gamma}
\qquad
  \nru{\tcase}
    {A_1 + A_2 \from \Gamma
      \qquad B \from \cext \Gamma A_1
      \qquad B \from \cext \Gamma A_2
    }{B \from \Gamma}
\qquad
  \nru{\tabort}
    {\tyzero \from \Gamma}
    {B \from \Gamma}
\end{gather*}
The skilled eye of the reader will immediately recognize the proof
rules of intuitionistic propositional logic (IPL) under the
Curry-Howard isomorphism.  A term such as
\(
  \tabs\,(\tabs\,(\tpair\,
    (\tvar\,(\tsuc\,\tzero))\,
    (\tabs\,(\tapp\,(\tvar\,(\tsuc\,\tzero))\,(\tvar\,\tzero)))))
\)
could in concrete syntax be rendered as
\(
  \lambda x.\,\lambda y.\,(x,\,\lambda z.\,y\,z)
  .
\)
We leave the exact connection to a printable
syntax of the STLC to imagination of the reader,
as we shall not be concerned with considering
concrete terms in this article.

@@

A direct extension of our presheaf semantics to positive types cannot
work.  For instance, with $\den \tyzero = \hat0$, simply
$\up[\tyzero] \tyzero (\tvar\,\tzero) : \bzero$ would give us an
inhabitant of the empty set, which means that reflection at the empty
type would not be definable.  Similarily, the setting
$\den{A + B} = \den A \hatplus \den B$ is refuted by
% $\up[\tyone + \tyone]^{\tyone + \tyone} (\tvar\,\tzero) : \bone \bplus \bone$
$\up[A + B]{A + B} (\tvar\,\tzero) : \den A (A{+}B) \bplus \den B (A{+}B)$
which would require us to make a decision of whether $A$ holds or $B$ holds
while only be given a hypothesis of type $A + B$.
Not even the usual interpretation of base types $\den o = \tNe\,o$
works in the presence of sums, as we would not be able to interpret
the term
$\tabs\,(\tcase\,(\tvar\,\tzero)\,(\tvar\,\tzero)\,(\tvar\,\tzero)) :
(o+o) \To o$ in our semantics, as $\Ne o (o+o)$ is empty.
What is needed are case distinctions on neutrals in the semantics,
allowing us the elimination of positive hypotheses before producing a
semantic value,
and we shall capture this capability in a strong monad $\C$
which can \emph{cover} the cases.

@@

    \tabs\,(\tapp\,(\tren\,\twk\,t)\,(\tvar\,\tzero))}

@@

      \,(\injP 1 {\tvar\,\tzero})
      \,(\injP 2 {\tvar\,\tzero})

@@

Assuming a family $\T\,N\,\Gamma$ of terms of negative type $N$ in
context $\Gamma$, values \fbox{$v : \PValT P \Gamma$}
of positive type $P$ are constructed by the
following rules:
\begin{gather*}
  \nru{\pvar}{P \in \Gamma}{\PValT P \Gamma}
\qquad
  \nru{\tthunk}{\T\,N\,\Gamma}{\PValTP{\Thunk N}\Gamma}
\\[2ex]
  \nru{\punit}{}{\PValT{\tyone}\Gamma}
\qquad
  \nru{\ppair}
      {\PValT{P_1}\Gamma \qquad \PValT{P_2}\Gamma}
      {\PValTP{P_1 \times P_2}\Gamma}
\qquad
  \nru{\tinj_i}{\PValT{P_i}\Gamma}{\PValTP{P_1 + P_2}\Gamma}
\end{gather*}

It relies on the previous definitions
with instance $\T = \tNf$.
\fbox{$\Tm N \Gamma$}:
\begin{gather*}
  \nru{\tret}
      {\PVal P \Gamma}
      {\TmP {\Comp P} \Gamma}
\quad
  \nru{\tabs}
      {\Tm N {(\cext \Gamma P)}}
      {\TmP {P \To N} \Gamma}
\quad
  \nru{\npair}
      {\Tm {N_1} \Gamma \qquad \Tm {N_2} \Gamma}
      {\TmP {N_1 \ntimes N_2} \Gamma}
\quad
  \nru{\nunit}
      {}
      {\Tm \top \Gamma}
\\[2ex]
  \nru{\tforce}
      {\PValP {\Thunk N} \Gamma}
      {\Tm N \Gamma}
\qquad
  \nru{\tapp}
      {\TmP {P \To N} \Gamma \qquad \PVal P \Gamma}
      {\Tm N \Gamma}
\qquad
  \nru{\tprj_i}
      {\TmP {N_1 \ntimes N_2} \Gamma}
      {\Tm {N_i} \Gamma}
\\[2ex]
% \end{gather*}
% \begin{gather*}
  \nru{\tbind}
      {\TmP {\Comp P} \Gamma \qquad \Tm N {(\cext \Gamma P)}}
      {\Tm N \Gamma}
\qquad
  \nru{\tsplit}
    {\TmP {P_1 \times P_2} \Gamma
      \qquad \Tm N {(\cext {\cext \Gamma P_1} P_2)}
    }{\Tm N \Gamma}
\\[2ex]
  \nru{\tcase}
    {\TmP {P_1 + P_2} \Gamma
      \qquad \Tm N {(\cext \Gamma P_1)}
      \qquad \Tm N {(\cext \Gamma P_2)}
    }{\Tm N \Gamma}
\qquad
  \nru{\tabort}
    {\Tm \tyzero \Gamma}
    {\Tm N \Gamma}
\end{gather*}

@@

\fbox{$\Ne N \Gamma$}:
\begin{gather*}
  \nru{\nvar}{\Thunk N \in \Gamma}{\Ne N \Gamma}
\qquad
  \nru{\tprj_i}
      {\NeP {N_1 \ntimes N_2} \Gamma}
      {\Ne {N_i} \Gamma}
\qquad
  \nru{\tapp}
      {\NeP {P \To N} \Gamma \qquad \PVal P \Gamma}
      {\Ne N \Gamma}
\end{gather*}
Pattern matching on variables is organized in the
inductively defined strong monad \fbox{$\Cov$}.
\begin{gather*}
  \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
\qquad
  \nru{\tbind}
      {\NeP {\Comp P} \Gamma \qquad \Cover \J (\cext \Gamma P)}
      {\Cover \J \Gamma}
\\[2ex]
  \nru{\tsplit}
      {P_1 \times P_2 \in \Gamma \qquad
       \Cover \J (\cext {\cext \Gamma {P_1}} {P_2})
      }
      {\Cover \J \Gamma}
\\[2ex]
  \nru{\tcase}
      {P_1 + P_2 \in \Gamma
       \qquad \Cover \J (\cext \Gamma {P_1})
       \qquad \Cover \J (\cext \Gamma {P_2})
      }
      {\Cover \J \Gamma}
\qquad
  \nru{\tabort}
      {\tyempty \in \Gamma}
      {\Cover \J \Gamma}
\end{gather*}
Finally, normal forms of negative types are defined as inductive
family \fbox{$\Nf N \Gamma$}.
\begin{gather*}
  \nru{\tne}
      {\CoverP {\tNe\,\natom} \Gamma}
      {\Nf {\natom} \Gamma}
\quad
  \nru{\tret}
      {\CoverP {\tVal\,P}  \Gamma}
      {\NfP {\Comp P} \Gamma}
\quad
  \nru{\npair}
      {\Nf {N_1} \Gamma \qquad \Nf {N_2} \Gamma}
      {\NfP {N_1 \ntimes N_2} \Gamma}
\quad
  \nru{\tabs}
      {\Nf N {\cext \Gamma P}}
      {\NfP {P \To N} \Gamma}
\end{gather*}
We again can run the cover monad on normal forms, \ie, have
$\trunNf : \Cov\,(\tNf N) \todot \tNf N$.

Reification $\down P : \den P \todot \tVal\,P$ at positive types produces
a normal value, and $\down N : \den P \todot \tNf\,N$ at negative types a
normal term.  During reification of function types $P \To N$ in
context $\Gamma$ we need to embed a fresh variable
$x : P \in \cext \Gamma P$ into $\den P$, breaking down $P$ to
positive atoms $\patom$ and negative remainders $\Thunk N$.
However, in $\den P$ we do not have case analysis available, thus,
positive reflection $\upG P : P \in \Gamma \to \Cover {\den P} \Gamma$
needs to run in the monad.  Luckily, as $\down N$ produces a normal
form, monadic intermediate computations are permitted under a final
$\trunNf$.  Negative reflection $\up N : \tNe\,N \todot \den N$ is as
before generalized from variables to neutrals, to handle the breaking
down of $N$ via eliminations.
\[
\begin{array}{lcl}
  \upG P & : & P \in \Gamma \to \den P \\
  \upG {\patom} \; x & = & x \\
  \upG {\tyone} \; x & = & \treturn\;\ttempty \\
  \upG {P_1 \times P_2} \; x & = & \tsplit\;x\;\tzip
\end{array}
\]

>>>>

Simple types shall be distinguished into positive types $P$ and
negative types $N$, depending on their root type former.  Function
($\To$) and product types ($\times$ and $\tyone$) are negative,
while base types ($\tybase$) and sum types ($+$ and $\tyzero$) are positive.
\[
\begin{array}{lrl@{\qquad}l}
  A,B,C & ::= & P \mid N & \mbox{simple types} \\
  P     & ::= & \tyzero \mid A + B \mid \tybase & \mbox{positive types} \\
  N     & ::= & \tyone \mid A \times B \mid A \To B
  % P     & ::= & a \mid A + B \mid \tyzero & \mbox{positive types} \\
  % N     & ::= & A \To B \mid A \times B \mid \tyone
    & \mbox{negative types} \\
\end{array}
\]
% Simple types shall be given by the grammar
% % \fbox{$S,T ::= P \mid S \To T \mid S \times T \mid \tyone \mid S + T
% %   \mid \tyzero$}.
% \fbox{$A,B,C ::= a \mid A \To B \mid A \times B \mid \tyone \mid A + B
%   \mid \tyzero$}.
% Herein, $a$ ranges over atomic types aka base types.
Intrinsically well-typed lambda-terms, in abstract syntax,
are just inhabitants $t$ of the indexed set \fbox{$A \from \Gamma$},
inductively defined by the following rules.
\begin{gather*}
  \nru{\tvar}{A \in \Gamma}{A \from \Gamma}
\qquad
  \nru{\tabs}{B \from \cext \Gamma A}{A \To B \from \Gamma}
\qquad
  \nru{\tapp}
    {A \To B \from \Gamma \qquad A \from \Gamma}
    {B \from \Gamma}
\\[2ex]
  \nru{\tunit}{}{\tyone \from \Gamma}
\qquad
  \nru{\tpair}
    {A_1 \from \Gamma \qquad A_2 \from \Gamma}
    {A_1 \times A_2 \from \Gamma}
\qquad
  \nru{\tprj_i}
    {A_1 \times A_2 \from \Gamma}
    {A_i \from \Gamma}
\\[2ex]
  \nru{\tinj_i}
    {A_i \from \Gamma}
    {A_1 + A_2 \from \Gamma}
\qquad
  \nru{\tcase}
    {A_1 + A_2 \from \Gamma
      \qquad B \from \cext \Gamma A_1
      \qquad B \from \cext \Gamma A_2
    }{B \from \Gamma}
\qquad
  \nru{\tabort}
    {\tyzero \from \Gamma}
    {B \from \Gamma}
\end{gather*}
The skilled eye of the reader will immediately recognize the proof
rules of intuitionistic propositional logic (IPL) under the
Curry-Howard isomorphism.  Using shorthand
\fbox{$\vv n = \tvar\,(\tsuc^n\,\tzero)$} for the $n$th variable,
a term such as
\(
  \tabs\,(\tabs\,(\tpair\,
    \vv1\,
    (\tabs\,(\tapp\,\vv1\,\vv0))))
\)
could in concrete syntax be rendered as
\(
  \lambda x.\,\lambda y.\,(x,\,\lambda z.\,y\,z)
  .
\)
We leave the exact connection to a printable
syntax of the STLC to imagination of the reader,
as we shall not be concerned with considering
concrete terms in this article.

@@

A direct extension of our presheaf semantics to positive types cannot
work.  For instance, with $\den \tyzero = \hat0$, simply
$\up[\tyzero] \tyzero \vv0 : \bzero$ would give us an
inhabitant of the empty set, which means that reflection at the empty
type would not be definable.  Similarily, the setting
$\den{A + B} = \den A \hatplus \den B$ is refuted by
% $\up[\tyone + \tyone]^{\tyone + \tyone} \vv0 : \bone \bplus \bone$
$\up[A + B]{A + B} \vv0 : \den A (A{+}B) \bplus \den B (A{+}B)$
which would require us to make a decision of whether $A$ holds or $B$ holds
while only be given a hypothesis of type $A + B$.
Not even the usual interpretation of base types $\den o = \tNe\,o$
works in the presence of sums, as we would not be able to interpret
the term
$\tabs\,(\tcase\,\vv0\,\vv0\,\vv0) :
(o+o) \To o$ in our semantics, as $\Ne o (o+o)$ is empty.
What is needed are case distinctions on neutrals in the semantics,
allowing us the elimination of positive hypotheses before producing a
semantic value,
and we shall capture this capability in a strong monad $\C$
which can \emph{cover} the cases.

@@

    \tabs\,(\tapp\,(\tren\,\twk\,t)\,\vv0)}

@@

      \,(\inj 1 {\vv0})
      \,(\inj 2 {\vv0})

@@

Assuming a family $\T\,N\,\Gamma$ of terms of negative type $N$ in
context $\Gamma$, values \fbox{$v : \PValT P \Gamma$}
of positive type $P$ are constructed by the
following rules:
\begin{gather*}
  \nru{\tvar}{P \in \Gamma}{\PValT P \Gamma}
  % \nru{\pvar}{P \in \Gamma}{\PValT P \Gamma}
\qquad
  \nru{\tthunk}{\T\,N\,\Gamma}{\PValTP{\Thunk N}\Gamma}
\\[2ex]
  \nru{\punit}{}{\PValT{\tyone}\Gamma}
\qquad
  \nru{\ppair}
      {\PValT{P_1}\Gamma \qquad \PValT{P_2}\Gamma}
      {\PValTP{P_1 \times P_2}\Gamma}
\qquad
  \nru{\tinj_i}{\PValT{P_i}\Gamma}{\PValTP{P_1 + P_2}\Gamma}
\end{gather*}

It relies on the previous definitions
with instance $\T = \tNf$.
\fbox{$\Tm N \Gamma$}:
\begin{gather*}
  \nru{\tret}
      {\PVal P \Gamma}
      {\TmP {\Comp P} \Gamma}
\quad
  \nru{\tabs}
      {\Tm N {(\cext \Gamma P)}}
      {\TmP {P \To N} \Gamma}
\quad
  \nru{\npair}
      {\Tm {N_1} \Gamma \qquad \Tm {N_2} \Gamma}
      {\TmP {N_1 \ntimes N_2} \Gamma}
\quad
  \nru{\nunit}
      {}
      {\Tm \top \Gamma}
\\[2ex]
  \nru{\tforce}
      {\PValP {\Thunk N} \Gamma}
      {\Tm N \Gamma}
\qquad
  \nru{\tapp}
      {\TmP {P \To N} \Gamma \qquad \PVal P \Gamma}
      {\Tm N \Gamma}
\qquad
  \nru{\tprj_i}
      {\TmP {N_1 \ntimes N_2} \Gamma}
      {\Tm {N_i} \Gamma}
\\[2ex]
% \end{gather*}
% \begin{gather*}
  \nru{\tbind}
      {\TmP {\Comp P} \Gamma \qquad \Tm N {(\cext \Gamma P)}}
      {\Tm N \Gamma}
\qquad
  \nru{\tsplit}
    {\PValP {P_1 \times P_2} \Gamma
      \qquad \Tm N {(\cext {\cext \Gamma P_1} P_2)}
    }{\Tm N \Gamma}
\\[2ex]
  \nru{\tcase}
    {\PValP {P_1 + P_2} \Gamma
      \qquad \Tm N {(\cext \Gamma P_1)}
      \qquad \Tm N {(\cext \Gamma P_2)}
    }{\Tm N \Gamma}
\qquad
  \nru{\tabort}
    {\PVal \tyzero \Gamma}
    {\Tm N \Gamma}
\end{gather*}

@@

\fbox{$\Ne N \Gamma$}:
\begin{gather*}
  \nru{\tforce}{\Thunk N \in \Gamma}{\Ne N \Gamma}
  % \nru{\nvar}{\Thunk N \in \Gamma}{\Ne N \Gamma}
\qquad
  \nru{\tprj_i}
      {\NeP {N_1 \ntimes N_2} \Gamma}
      {\Ne {N_i} \Gamma}
\qquad
  \nru{\tapp}
      {\NeP {P \To N} \Gamma \qquad \PVal P \Gamma}
      {\Ne N \Gamma}
\end{gather*}
Pattern matching on variables is organized in the
inductively defined strong monad \fbox{$\Cov$}.
\begin{gather*}
  \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
\qquad
  \nru{\tbind}
      {\NeP {\Comp P} \Gamma \qquad \Cover \J (\cext \Gamma P)}
      {\Cover \J \Gamma}
\\[2ex]
  \nru{\tsplit}
      {P_1 \times P_2 \in \Gamma \qquad
       \Cover \J (\cext {\cext \Gamma {P_1}} {P_2})
      }
      {\Cover \J \Gamma}
\\[2ex]
  \nru{\tcase}
      {P_1 + P_2 \in \Gamma
       \qquad \Cover \J (\cext \Gamma {P_1})
       \qquad \Cover \J (\cext \Gamma {P_2})
      }
      {\Cover \J \Gamma}
\qquad
  \nru{\tabort}
      {\tyempty \in \Gamma}
      {\Cover \J \Gamma}
\end{gather*}
Finally, normal forms of negative types are defined as inductive
family \fbox{$\Nf N \Gamma$}.
\begin{gather*}
  \nru{\tne}
      {\CoverP {\tNe\,\natom} \Gamma}
      {\Nf {\natom} \Gamma}
\quad
  \nru{\tret}
      {\CoverP {\tVal\,P}  \Gamma}
      {\NfP {\Comp P} \Gamma}
\quad
  \nru{\npair}
      {\Nf {N_1} \Gamma \qquad \Nf {N_2} \Gamma}
      {\NfP {N_1 \ntimes N_2} \Gamma}
\quad
  \nru{\tabs}
      {\Nf N {\cext \Gamma P}}
      {\NfP {P \To N} \Gamma}
\end{gather*}
We again can run the cover monad on normal forms, \ie, have
$\trunNf : \Cov\,(\tNf N) \todot \tNf N$.

Reification $\down P : \den P \todot \tVal\,P$ at positive types produces
a normal value, and $\down N : \den P \todot \tNf\,N$ at negative types a
normal term.  During reification of function types $P \To N$ in
context $\Gamma$ we need to embed a fresh variable
$x : P \in \cext \Gamma P$ into $\den P$, breaking down $P$ to
positive atoms $\patom$ and negative remainders $\Thunk N$.
However, in $\den P$ we do not have case analysis available, thus,
positive reflection $\upG P : P \in \Gamma \to \Cover {\den P} \Gamma$
needs to run in the monad.  Luckily, as $\down N$ produces a normal
form, monadic intermediate computations are permitted under a final
$\trunNf$.  Negative reflection $\up N : \tNe\,N \todot \den N$ is as
before generalized from variables to neutrals, to handle the breaking
down of $N$ via eliminations.
%
In the following definition of reflection we use the abbreviation
\fbox{$\tfresh^P_\Gamma = \up[\cext \Gamma P] P \vv0 : \Cov\,\den P$}.
\[
\def\arraystretch{1.3}
\begin{array}{l@{\,}l@{~~}c@{~~}l}
  \multicolumn 2 l {\upG P} & : & P \in \Gamma \to \Cov\,\den P \, \Gamma\\
  \upG {\patom}         & x & = & x \\
  \upG {\tyone}         & x & = & \treturn\;\ttempty \\
  \upG {P_1 \times P_2} & x & = & \tsplit\;x\,
    \left(
     (\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_1}\,\vv1)
     \star
     (\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_2}\,\vv0)
    \right) \\
    % (\tzip
    % \,(\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_1}\,\vv1)
    % \,(\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_2}\,\vv0)
    % ) \\
  \upG{\tyzero}         & x & = & \tabort\;x \\
  \upG{P_1 + P_2}       & x & = & \tcase\;x
    \; (\tmap\;\iota_1\;\tfresh^{P_1}_\Gamma)
    \; (\tmap\;\iota_2\;\tfresh^{P_2}_\Gamma)
    \\
  \upG{\Thunk N} & x & = & \treturn\,(\upG N (\tforce\,x)) \\
\end{array}
\,
\begin{array}{lcl}
  \down P & : & \den P \todot \tVal\,P \\
  \down {\patom} & = & \tvar \\
  \downG {\tyone} () & = & \punit \\
  \downG {P_1 \times P_2}(a_1,a_2) & = & \ppair
    \,(\downG {P_1} a_1)
    \,(\downG {P_2} a_2)
    \\
  \down {\tyzero} & = & \tmagic \\
  \down {P_1 + P_2} & = &
    [  \tinj_1 \comp \down{P_1}
    ,\ \tinj_2 \comp \down{P_2}
    ] \\
  \down {\Thunk N} & = & \tthunk \circ \down N \\
\end{array}
\]
Reflection at positive pairs uses monoidal functoriality
$\C\,\A \todot \C\,\B \todot \C\,(\A \hattimes \B)$
called $\star$ by McBride and Paterson \cite[Section 7]{mcbridePaterson:applicative}
which for monads $\C$ can be defined by
$c \star d = \tjoin \; (\tmap\;(\lambda\,b.\ (c,\, \treturn\,b))\;d$.
% $\tzip\;c = \tjoin \comp \tmap\,(\lambda\,a.\ (c,\, \treturn\,a)$.

For negative product types, reflection and reification works as before:
\[
\def\arraystretch{1.3}
\begin{array}{l@{\,}l@{~~}c@{~~}l}
  \multicolumn 2 l {\up N} & : & \tNe\,N \todot \den N \\
  \upG \natom & u & = & \treturn\,u \\
  \upG \top   & u & = & \ttempty \\
  \upG {N_1 \ntimes N_2} & u & = & \left(
    \upG {N_1} \; (\prj 1 u)
    ,\
    \upG {N_2} \; (\prj 2 u)
    \right) \\
\end{array}
\
\begin{array}{l@{\,}l@{~~}c@{~~}l}
  \multicolumn 2 l {\down N} & : & \den N \todot \tNf\;N \\
  \downG \natom & c  & = & \tne\,c \\
  \downG \top   & () & = & \nunit \\
  \downG {N_1 \ntimes N_2} & (b_1,b_2) & = & \npair
    \, (\downG {N_1} \; b_1)
    \, (\downG {N_2} \; b_1)
    \\
\end{array}
\]
Reflection for function types is also unchanged, except that $\tapp$
expects a value argument now.
\[
\def\arraystretch{1.3}
\begin{array}{l@{\,}l@{~~}c@{~~}l}
  \upG {P \To N} & u & = & \lambda
    \;(\tau : \Gamma \sublist \Delta)
    \;(a : \den P \Delta)
    . \
    \up[\Delta] N \left( \tapp\,(\tren\,\tau\,u)\,(\down[\Delta] P a) \right)
    \\
  \downG {P \To N} & f & = & \tabs \left( \trunNf \left(
    \stmap
      \left(
        \lambda
          \;(\tau : (\cext \Gamma P) \sublist \Delta)
          \;(a : \den P \Delta)
          . \
          f\; (\twk^P \rcomp \tau)\; a
        \right)
      \,\tfresh^P_\Gamma
    \right) \right)
    \\
\end{array}
\]
